Free Theorems for Nested Types

### 2020-06-24 free-theorems-nesteds/paper4.tex
added paper4

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\gamma^0},\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\ol{\gamma^0} \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[3ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[2ex]
\AXC{$\Gamma;\ol{\gamma^0},\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi,\ol{\gamma^0} \vdash G}$}
\BIC{$\Gamma;\Phi,\ol{\gamma^0} \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F)\,\ol{G}$} 
\DisplayProof
\end{array}\]

@@

Definition~\ref{def:wftypes} allows the formation of all of the
(closed) nested types from the introduction:
\[\begin{array}{lll}
\mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
\beta\; = \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
\beta)\,\alpha\\ 
\mathit{PTree}\,\alpha & = & (\mu \phi. \lambda \beta.\,\beta +
\phi\,(\beta \times \beta))\,\alpha\\
\mathit{Forest}\,\alpha & = & (\mu \phi. \lambda \beta. \,\onet +
\beta \times \mathit{PTree}\,(\phi \beta))\,\alpha\\ 
\mathit{Bush}\,\alpha & = & (\mu \phi.\lambda \beta. \,\onet + \beta
\times \phi\,(\phi\beta))\,\alpha 
\end{array}\]
Each of these types can be considered either functorial in $\alpha$ or
not, according to whether $\alpha$ is in the functorial or
non-functorial context in which it is formed. For example, if
$\emptyset;\alpha \vdash \mathit{List} \,\alpha$, then $\vdash
\Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is well-formed; if
$\alpha;\emptyset \vdash \List \,\alpha$, then it is not.  If
$\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
\gamma \times \beta$, then both $\gamma;\emptyset \vdash \Nat^\alpha
(\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ and
$\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\, \alpha) \,
(\mathit{Tree} \, \alpha\, \gamma)$ are well-formed.  Although types
can be functorial in variables of arity greater than $0$, the body $F$
of a type $(\mu \phi.\lambda \ol\alpha.F)\,\ol G$ cannot. For example,
the type $\mathit{GRose}\, \phi \, \alpha = \mu\beta . \alpha \times
\phi \beta$ can be functorial or not in $\alpha$, but cannot be
functorial in $\phi$.  As a result, $\mathit{GRose}\,\phi\,\alpha$
cannot be the (co)domain of a $\Nat$ type representing a natural
transformation functorial in $\phi$, and the type $\mu \phi. \lambda
\alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
system. These restrictions on higher-arity functoriality are necessary
for constructing our parametric model.  Fortunately, they are entirely
natural, and {\color{blue} preclude only truly bizarre nested types}.
%do not preclude any nested
%types {\color{blue} occurring in practice}.

%If $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
%\gamma \times \beta$, then Definition~\ref{def:wftypes} also allows
%the derivation of, e.g., the type $\gamma; \emptyset \vdash
%\Nat^\alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
%\gamma)$ representing a natural transformation from lists to trees
%that is natural in $\alpha$ but not necessarily in $\gamma$. We
%emphasize that types can be functorial in variables of arity greater
%than $0$. For example, the type $\mathit{GRose}\, \phi \, \alpha =
%\mu\beta . \alpha \times \phi \beta$ can be functorial in $\phi$ if
%$\phi \in \Phi$. As usual, whether $\phi \in \Gamma$ or $\phi \in
%\Phi$ determines whether types such as $\Nat^\alpha ( \mathit{GRose}
%\, \phi \, \alpha)\, (List\, \alpha)$ are well-formed. But even if
%$\mathit{GRose}$ is functorial in $\phi$, it still cannot be the
%domain of a $\Nat$ type representing a natural transformation in
%$\phi$. This is because our calculus does not allow naturality in
%variables of arity greater than $0$.

@@

\[\begin{array}{c}
%\AXC{$\Gamma; \emptyset \vdash \Nat^{\ol{\alpha}} \,F\,G : \T$}
\AXC{$\Gamma; \ol{\alpha} \vdash F$}
\AXC{$\Gamma; \ol\gamma, \ol{\alpha} \vdash G$}
\AXC{$\Gamma; \ol\gamma, \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
\TIC{$\Gamma; \ol\gamma
  %\emptyset
  \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
\DisplayProof
\\\\
\AXC{$\Gamma; \ol\gamma
  %\emptyset
  \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
\AXC{$\ol{\Gamma;\Phi,\ol\gamma \vdash K}$}
\AXC{$\Gamma;\Phi,\ol\gamma \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
\TIC{$\Gamma;\Phi,\ol\gamma \,|\, \Delta \vdash t_{\ol K} s:
  G[\overline{\alpha := K}]$}
\DisplayProof
\\\\
\AXC{$\Gamma; \ol{\phi}, \ol{\gamma} \vdash H$}
\AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
\AXC{$\ol{\Gamma; \ol\alpha, \ol{\beta},\ol{\gamma} \vdash
    G}$\hspace*{0.3in}$\ol{\Gamma;\ol\alpha~|~\Delta \vdash t :
  \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}$}
\TIC{$\Gamma; \ol\alpha
  %\emptyset
  ~|~\Delta
%  \ol{ x : \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}
  %  \emptyset
  \vdash \map^{\ol{F},\ol{G}}_H \, \ol t :
%  \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
  \Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
      :=_{\ol{\beta}} G}]$} 
\DisplayProof
\\\\
%\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : H[\phi := (\mu \phi.\lambda 
%  \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \tau}]$}
%\AXC{$\ol{\Gamma;\Phi \vdash \tau : \F}$}
\AXC{$\Gamma; \phi, \ol{\alpha},\ol{\gamma} \vdash H$}
\UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
  \Nat^{\ol{\beta},\ol{\gamma}} H[\phi :=_{\ol{\beta}} (\mu
    \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
  \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
\DisplayProof
\\\\
\AXC{$\Gamma; \phi,\ol{\alpha}, \ol{\gamma} \vdash H$}
\AXC{$\Gamma; \ol{\beta}, \ol{\gamma} \vdash F$}
%\AXC{$\Gamma;\emptyset \,|\, \Delta \vdash t :
%  \Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha
%      := \beta}]\,F$} 
%\TIC{$\Gamma;\emptyset \,|\, \Delta \vdash
%  \fold_{H,F}\, t : \Nat^{\ol{\beta}, \ol{\gamma}}\,((\mu
%  \phi.\lambda \ol{\alpha}.H)\ol{\beta})\,F$} 
\BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
  \Nat^\emptyset\; (\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\ol{\beta},
    \ol{\gamma}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta})\,F)$}
\DisplayProof
\end{array}\]
\end{dfn}

@@

  \alpha} t & : & \Nat^{\alpha} (\mathit{PTree}\,(\alpha \times
\alpha))\, (\mathit{PTree}\,(\alpha \times \alpha))

@@

for $t : \Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times
\alpha)$, 

@@

    {\color{blue} \nat}))$, all of whose argument types are

@@

    for nested types, even when those nested types are semantically

@@

arguments are parameterized over disjoint sets of type variables. The
return type of the function can be parameterized over any or all of
the type variables occurring in its argument types.  For example, we
can represent the type of a zip-like function for bushes.

@@

Note, however, that even some recursive functions of a single proper

@@

fixpoints studied here.

@@

The presence of the ``extra'' functorial variables in $\ol{\gamma}$ in
the rules for $\map^{\ol{F},\ol{G}}_H$, $\tin_H$, and $\fold^F_H$
merit special mention. They allow us to map or fold polymorphic
functions over nested types. Consider, for example, the polymorphic
function $\mathit{flatten} : \Nat^\beta
(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ that uniformly maps
perfect trees to lists. Even in the absence of extra variables the
instance of $\map$ required to map each non-functorial monomorphic
instantiation of $\mathit{flatten}$ over a list of perfect trees is
well-typed:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$ \hspace*{0.3in}
  $\Gamma;\emptyset~|~\Delta \vdash t
  : \Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\,
  G)$}
\TIC{$\Gamma;\emptyset \,|\, \Delta \vdash \map^{\mathit{PTree}\,F,
    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} \,t :
  \Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
  F))\,(\mathit{List}\, (\mathit{List}\, G))$} \DisplayProof
\end{array}\]
But in the absence of $\ol \gamma$, if $\Gamma;
\emptyset~|~\Delta\vdash t : \Nat^\beta(\mathit{PTree}\,
\beta)\,(\mathit{List}\, \beta)$ then the instance
\[\Gamma;\emptyset~|~\Delta \vdash
\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}\,t
: \Nat^\beta\,(\mathit{List}\, (\mathit{PTree}\,
\beta))\,(\mathit{List}\, (\mathit{List}\, \beta))\]
%\[\Gamma;\emptyset \,|\, x : \Nat^\beta(\mathit{PTree}\,
%\beta)\,(\mathit{List}\, \beta) \vdash
%\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}\,x
%: \Nat^\beta\,(\mathit{List}\,
%(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
%\beta))\] of $\mathtt{map}$
required to map the {\em polymorphic}
$\mathit{flatten}$ function over a list of perfect trees is not: in
that setting the functorial contexts for $F$ and $G$ in the rule for
$\map^{F,G}_H$ would have to be empty, but the fact that the
polymorphic $\mathit{flatten}$ function is functorial in some
variable, say $\delta$, means that it cannot possibly have a type of
the form $\Nat^\emptyset F\, G$ that would be required for it to be
the function input to $\map$. Since untypeability of this instance of
$\map$ is unsatisfactory in a polymorphic calculus, where we naturally
expect to be able to manipulate entire polymorphic functions rather
than just their monomorphic instances, we use the ``extra'' variables
in $\ol \gamma$ to remedy the situation. Specifically, the rules from
Definition~\ref{def:well-formed-terms} ensure that the instance of
$\map$ needed to map the polymorphic $\mathit{flatten}$ function is
typeable as follows:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha,\beta \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\beta \vdash \mathit{PTree}\,\beta \hspace*{0.3in}
  \Gamma;\beta \vdash \mathit{List}\,\beta$\hspace*{0.3in}
$\Gamma;\emptyset~|~\Delta \vdash t : \Nat^\beta(\mathit{PTree}\,
  \beta)\,(\mathit{List}\, \beta)$}
\BIC{$\Gamma;\emptyset~|~\Delta \vdash
  \map^{F,G}_{\mathit{List}\,\alpha} \, t :
 \Nat^\beta\,(\mathit{List}\,
  (\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
  \beta))$}
\DisplayProof
\end{array}\]
Similar remarks explain the appearance of $\ol \gamma$ in the typing
rules for $\tin$ and $\fold$.

@@

\item If \,$\Gamma; \ol\gamma
  %\emptyset

@@

  $\setsem{\Gamma; \ol\gamma
    %\emptyset
    \vdash \Nat^{\ol{\alpha}}\,F\,G} f =$\\
%  \id_{\setsem{\Gamma; \Phi
      %\emptyset
%      \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho}$
\hspace*{0.2in} $\lambda t : \setsem{\Gamma;\ol\gamma \vdash
    \Nat^{\ol{\alpha}}\,F\,G}\rho.\, 
  (\lambda \ol{A}. \setsem{\Gamma; \ol\gamma,\ol\alpha \vdash G}f[\ol{\alpha
      := id_A}]) \circ t$  

@@

\dsem{\Gamma;\emptyset \,|\, \Delta,x :F \vdash x : F} \rho& = &

@@

\dsem{\Gamma;\ol\gamma \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
    \alpha} \,F \,G}\rho & = &  \curry (\dsem{\Gamma;\ol\gamma,\overline \alpha

@@

\dsem{\Gamma;\Phi,\ol\gamma \,|\, \Delta \vdash t_{\overline K} s:

@@

  \lambda d.\,(\dsem{\Gamma;\ol\gamma \,|\, \Delta \vdash t :
  \Nat^{\overline{\alpha}} \,F \,G}\rho\; d)_{\overline{\dsem{\Gamma;\Phi,\ol\gamma

@@

 & & \hspace*{0.5in} \dsem{\Gamma;\Phi,\ol\gamma \,|\,

@@

@@

\dsem{\Gamma;\ol\alpha \,|\, \Delta \vdash \map^{\ol{F},\ol{G}}_H
\,\ol t: \Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
      :=_{\ol{\beta}} G}]} \rho
& = & \lambda d\, \ol{C}.\,
\dsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}\\ 
\hspace*{0.79in}  & & \hspace*{0.1in}\id_{\rho[\ol{\gamma :=
      C}]}[\ol{\phi := \lambda
    \ol{B}. (\dsem{\Gamma;\ol\alpha\,|\,\Delta \vdash t : 
    \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}\rho\,d)_{\ol{B}\,\ol{C}}}]\\

@@

Straightforward calculation shows that terms of $\Nat$ type behave as
natural transformations with respect to their source and target
functorial types, i.e., we have

>>>>

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[3ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[2ex]
\AXC{$\Gamma;\ol{\gamma^0},\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi,\ol{\gamma^0} \vdash G}$}
\BIC{$\Gamma;\Phi,\ol{\gamma^0} \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F)\,\ol{G}$} 
\DisplayProof
\end{array}\]

@@

Definition~\ref{def:wftypes} allows the formation of all of the
(closed) nested types from the introduction:
\[\begin{array}{lll}
\mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
\beta\; = \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
\beta)\,\alpha\\ 
\mathit{PTree}\,\alpha & = & (\mu \phi. \lambda \beta.\,\beta +
\phi\,(\beta \times \beta))\,\alpha\\
\mathit{Forest}\,\alpha & = & (\mu \phi. \lambda \beta. \,\onet +
\beta \times \mathit{PTree}\,(\phi \beta))\,\alpha\\ 
\mathit{Bush}\,\alpha & = & (\mu \phi.\lambda \beta. \,\onet + \beta
\times \phi\,(\phi\beta))\,\alpha 
\end{array}\]
Each of these types can be considered either functorial in $\alpha$ or
not, according to whether $\alpha$ is in the functorial or
non-functorial context in which it is formed. For example, if
$\emptyset;\alpha \vdash \mathit{List} \,\alpha$, then $\vdash
\Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is well-formed; if
$\alpha;\emptyset \vdash \List \,\alpha$, then it is not.  If
$\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
\gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
(\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is
well-formed, but $\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\,
\alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is not.  Although types
can be functorial in variables of arity greater than $0$, the body $F$
of a type $(\mu \phi.\lambda \ol\alpha.F)\,\ol G$ cannot. For example,
the type $\mathit{GRose}\, \phi \, \alpha = \mu\beta . \alpha \times
\phi \beta$ can be functorial or not in $\alpha$, but cannot be
functorial in $\phi$. In particular, $\mathit{GRose}\,\phi\,\alpha$
cannot be the (co)domain of a $\Nat$ type representing a natural
transformation functorial in $\phi$, and the type $\mu \phi. \lambda
\alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
system. The restrictions imposed by our typing rules will ensure that
set interpretations of types are functors. In
Section~\ref{sec:conclusion} we discuss how to lessen the expressivity
constraints they impose.
%These restrictions on higher-arity functoriality are necessary
%for constructing our parametric model. Fortunately, they are entirely
%natural, and {\color{blue} preclude only truly bizarre nested types}.
%do not preclude any nested
%types {\color{blue} occurring in practice}.

%If $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
%\gamma \times \beta$, then Definition~\ref{def:wftypes} also allows
%the derivation of, e.g., the type $\gamma; \emptyset \vdash
%\Nat \alpha (\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\,
%\gamma)$ representing a natural transformation from lists to trees
%that is natural in $\alpha$ but not necessarily in $\gamma$. We
%emphasize that types can be functorial in variables of arity greater
%than $0$. For example, the type $\mathit{GRose}\, \phi \, \alpha =
%\mu\beta . \alpha \times \phi \beta$ can be functorial in $\phi$ if
%$\phi \in \Phi$. As usual, whether $\phi \in \Gamma$ or $\phi \in
%\Phi$ determines whether types such as $\Nat^\alpha ( \mathit{GRose}
%\, \phi \, \alpha)\, (List\, \alpha)$ are well-formed. But even if
%$\mathit{GRose}$ is functorial in $\phi$, it still cannot be the
%domain of a $\Nat$ type representing a natural transformation in
%$\phi$. This is because our calculus does not allow naturality in
%variables of arity greater than $0$.

@@

\[\begin{array}{c}
%\AXC{$\Gamma; \emptyset \vdash \Nat^{\ol{\alpha}} \,F\,G : \T$}
\AXC{$\Gamma; \ol{\alpha} \vdash F$}
\AXC{$\Gamma; \ol{\alpha} \vdash G$}
\AXC{$\Gamma; \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
\TIC{$\Gamma; \emptyset
  \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
\DisplayProof
\\\\
\AXC{$\Gamma; \emptyset
  \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
\AXC{$\ol{\Gamma;\Phi \vdash K}$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
\TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
  G[\overline{\alpha := K}]$}
\DisplayProof
\\\\
\AXC{$\Gamma; \ol{\phi}, \ol{\gamma} \vdash H$}
\AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
\AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash
    G}$}
\TIC{$\Gamma; \emptyset
  ~|~\emptyset
%  \ol{ x : \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}
  %  \emptyset
  \vdash \map^{\ol{F},\ol{G}}_H :
  \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
  (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
      :=_{\ol{\beta}} G}])$} 
\DisplayProof
\\\\
%\AXC{$\Gamma;\Phi \,|\, \Delta \vdash t : H[\phi := (\mu \phi.\lambda 
%  \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \tau}]$}
%\AXC{$\ol{\Gamma;\Phi \vdash \tau : \F}$}
\AXC{$\Gamma; \phi, \ol{\alpha},\ol{\gamma} \vdash H$}
\UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
  \Nat^{\ol{\beta},\ol{\gamma}} H[\phi :=_{\ol{\beta}} (\mu
    \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
  \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
\DisplayProof
\\\\
\AXC{$\Gamma; \phi,\ol{\alpha}, \ol{\gamma} \vdash H$}
\AXC{$\Gamma; \ol{\beta}, \ol{\gamma} \vdash F$}
%\AXC{$\Gamma;\emptyset \,|\, \Delta \vdash t :
%  \Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi :=_{\ol{\beta}} F][\ol{\alpha
%      := \beta}]\,F$} 
%\TIC{$\Gamma;\emptyset \,|\, \Delta \vdash
%  \fold_{H,F}\, t : \Nat^{\ol{\beta}, \ol{\gamma}}\,((\mu
%  \phi.\lambda \ol{\alpha}.H)\ol{\beta})\,F$} 
\BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
  \Nat^\emptyset\; (\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\ol{\beta},
    \ol{\gamma}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta})\,F)$}
\DisplayProof
\end{array}\]
\end{dfn}

@@

  \alpha} & : & \Nat^\emptyset
(\Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times
\alpha))\,
(\Nat^{\alpha} (\mathit{PTree}\,(\alpha \times
\alpha))\, (\mathit{PTree}\,(\alpha \times \alpha)))

@@

@@

    (\mu \beta. 1 + \beta)))$, all of whose argument types are

@@

    for nested types, even when they are semantically

@@

arguments are parameterized over disjoint sets of type variables; the
return type of such a function can be parameterized over only 
%any or all of
the type variables occurring in its final argument's type.
%For example, we can represent the type of a zip-like function for bushes.

@@

Moreover, even some recursive functions of a single proper

@@

fixpoints studied here. Enhancing the expressivity of our calculus is
discussed further in Section~\ref{sec:conclusion}.

@@

The presence of the ``extra'' functorial variables in $\ol{\gamma}$ in
the rules for $\map^{\ol{F},\ol{G}}_H$, $\tin_H$, and $\fold^F_H$
merit special mention. They allow us to map or fold polymorphic
functions over nested types. Consider, for example, the polymorphic
function $\mathit{flatten} : \Nat^\beta
(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ that uniformly maps
perfect trees to lists. Even in the absence of extra variables the
instance of $\map$ required to map each non-functorial monomorphic
instantiation of $\mathit{flatten}$ over a list of perfect trees is
well-typed:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
\end{array}\]
But in the absence of $\ol \gamma$, the instance
\[\Gamma;\emptyset~|~\emptyset \vdash
\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
(\mathit{PTree}\, 
\beta))\,(\mathit{List}\, (\mathit{List}\, \beta)))\]
%\[\Gamma;\emptyset \,|\, x : \Nat^\beta(\mathit{PTree}\,
%\beta)\,(\mathit{List}\, \beta) \vdash
%\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}\,x
%: \Nat^\beta\,(\mathit{List}\,
%(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
%\beta))\] of $\mathtt{map}$
required to map the {\em polymorphic}
$\mathit{flatten}$ function over a list of perfect trees is not: in
that setting the functorial contexts for $F$ and $G$ in the rule for
$\map^{F,G}_H$ would have to be empty, but the fact that the
polymorphic $\mathit{flatten}$ function is functorial in some
variable, say $\delta$, means that it cannot possibly have a type of
the form $\Nat^\emptyset F\, G$ that would be required for it to be
the function input to $\map$. Since untypeability of this instance of
$\map$ is unsatisfactory in a polymorphic calculus, where we naturally
expect to be able to manipulate entire polymorphic functions rather
than just their monomorphic instances, we use the ``extra'' variables
in $\ol \gamma$ to remedy the situation. Specifically, the rules from
Definition~\ref{def:well-formed-terms} ensure that the instance of
$\map$ needed to map the polymorphic $\mathit{flatten}$ function is
typeable as follows:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha,\beta \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\beta \vdash \mathit{PTree}\,\beta \hspace*{0.3in}
  \Gamma;\beta \vdash \mathit{List}\,\beta$}
\BIC{$\Gamma;\emptyset~|~\emptyset \vdash
  \map^{F,G}_{\mathit{List}\,\alpha}  : \Nat^\emptyset
  (\Nat^\beta(\mathit{PTree}\, 
  \beta)\,(\mathit{List}\, \beta))\,
 (\Nat^\beta\,(\mathit{List}\,
  (\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
  \beta)))$}
\DisplayProof
\end{array}\]
Similar remarks explain the appearance of $\ol \gamma$ in the typing
rules for $\tin$ and $\fold$.

@@

\item If \,$\Gamma; \emptyset

@@

  $\setsem{\Gamma; \emptyset
    \vdash \Nat^{\ol{\alpha}}\,F\,G} f =
  \id_{\setsem{\Gamma; \emptyset
      \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho}$
%\hspace*{0.2in} $\lambda t : \setsem{\Gamma;\emptyset\vdash
%    \Nat^{\ol{\alpha}}\,F\,G}\rho.\, 
%  (\lambda \ol{A}. \setsem{\Gamma; \ol\alpha \vdash G}f[\ol{\alpha
%      := id_A}]) \circ t$  

@@

@@

\dsem{\Gamma;\emptyset \,|\, \Delta \vdash L_{\overline \alpha} x.t : \Nat^{\overline
    \alpha} \,F \,G}\rho & = &  \curry (\dsem{\Gamma;\overline \alpha

@@

\dsem{\Gamma;\Phi \,|\, \Delta \vdash t_{\overline K} s:

@@

  \lambda d.\,(\dsem{\Gamma;\emptyset \,|\, \Delta \vdash t :
  \Nat^{\overline{\alpha}} \,F \,G}\rho\; d)_{\overline{\dsem{\Gamma;\Phi

@@

 & & \hspace*{0.5in} \dsem{\Gamma;\Phi \,|\,

@@

%\dsem{\Gamma;\Phi \,|\, \Delta,x :F \vdash x : F} \rho& = &
%\pi_{|\Delta|+1}\\

@@

\llbracket \Gamma;\emptyset \,|\, \emptyset \vdash \map^{\ol{F},\ol{G}}_H
  : \Nat^\emptyset (\Nat^{\ol\beta,\ol\gamma} F\,G)
& = & \lambda d\, \ol\eta\,\ol{C}.\,
\dsem{\Gamma; \ol{\phi},\ol{\gamma}\vdash H}\id_{\rho[\ol{\gamma:=
      C}]}[\ol{\phi := \lambda \ol B. \eta_{\ol B\, \ol C}}]\\ 
\hspace*{0.5in}
  (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\,H[\ol{\phi
      :=_{\ol{\beta}} G}]) \rrbracket^{\mathsf D} \rho & &\\
%\hspace*{0.79in}  & & \hspace*{0.1in}\id_{\rho[\ol{\gamma :=
%      C}]}[\ol{\phi := \lambda
%    \ol{B}. (\dsem{\Gamma;\ol\alpha\,|\,\Delta \vdash t : 
%    \Nat^{\ol{\beta},\ol{\gamma}}\,F\,G}\rho\,d)_{\ol{B}\,\ol{C}}}]\\

@@

Define, for $\Gamma; \ol{\alpha} \vdash F$, the term $\id_F$ to be
$\Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
\Nat^{\ol{\alpha}} F\,F$ and, for terms $\Gamma; \emptyset \,|\,
\Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \emptyset
\,|\, \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$, the {\em
  composition} $s \circ t$ of $t$ and $s$ to be $\Gamma;
\emptyset\,|\, \Delta \vdash L_{\overline{\alpha}}
x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
\Nat^{\overline{\alpha}} F\,H$. Then $\setsem{\Gamma; \emptyset \,|\,
  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
  [\ol{\alpha := A}]}$ for any set environment $\rho$, and
$\setsem{\Gamma; \emptyset \,|\, \Delta \vdash s \circ t:
  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \emptyset \,|\,
  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
